---
title: 'Bios 6301: Assignment 3'
author: "Alexander Thiemicke"
email: alexander.thiemicke@vanderbilt.edu
date: "2015-10-09"
#"'r format(Sys.time(), '%B, %d, %Y')"
output: pdf_document
---



50 points total.





### Question 1 ###

**10 points**

1. Use GitHub to turn in the first three homework assignments. Make sure the teacher (couthcommander) and TA (trippcm) are collaborators. (5 points)

1. Commit each assignment individually.  This means your repository should have at least three commits. (5 points)

### Question 2 ###

**15 points**

Write a simulation to calculate the power for the following study
design.  The study has two variables, treatment group and outcome.
There are two treatment groups (0, 1) and they should be assigned
randomly with equal probability.  The outcome should be a random normal
variable with a mean of 60 and standard deviation of 20.  If a patient
is in the treatment group, add 5 to the outcome.  5 is the true
treatment effect.  Create a linear of model for the outcome by the
treatment group, and extract the p-value (hint: see assigment1).
Test if the p-value is less than or equal to the alpha level, which
should be set to 0.05.

Repeat this procedure 1000 times. The power is calculated by finding
the percentage of times the p-value is less than or equal to the alpha
level.  Use the `set.seed` command so that the professor can reproduce
your results.

1. Find the power when the sample size is 100 patients. (10 points)

```{r}
set.seed(63)
outcome1 <- rnorm(50, mean=62.5, sd=20)
outcome2 <- rnorm(50, mean=57.5, sd=20)
outcome = c(outcome1, outcome2)
treatment <- c(rep.int( 1, 50), rep.int(0, 50))

y = data.frame(treatment, c(outcome1, outcome2))
mod3 = lm(outcome ~ treatment, dat=y)
summary(mod3)
coef(summary(mod3))
alpha = 0.05
bu1 <- 0
for (i in 1:1000) {
  outcome1 <- rnorm(50, mean=62.5, sd=20)
outcome2 <- rnorm(50, mean=57.5, sd=20)
outcome = c(outcome1, outcome2)
treatment <- c(rep.int( 1, 50), rep.int(0, 50))

y = data.frame(treatment, c(outcome1, outcome2))
mod3 = lm(outcome ~ treatment, dat=y)
summary(mod3)
    i <- coef(summary(mod3))
    p <- (pt(i[2,3], 99, lower.tail = FALSE, log.p = FALSE) *2)
    if (p < alpha) 
      bu1 = bu1 + 1
  
}
power=bu1/1000
power

```

The test has a power of 21.8%

> set.seed(63)
> outcome1 <- rnorm(50, mean=62.5, sd=20)
> outcome2 <- rnorm(50, mean=57.5, sd=20)
> outcome = c(outcome1, outcome2)
> treatment <- c(rep.int( 1, 50), rep.int(0, 50))
> 
> y = data.frame(treatment, c(outcome1, outcome2))
> mod3 = lm(outcome ~ treatment, dat=y)
> summary(mod3)

Call:
lm(formula = outcome ~ treatment, data = y)

Residuals:
    Min      1Q  Median      3Q     Max 
-61.772 -13.499   2.687  14.130  38.787 

Coefficients:
            Estimate Std. Error t value Pr(>|t|)    
(Intercept)   54.995      2.822   19.49   <2e-16 ***
treatment      8.781      3.991    2.20   0.0302 *  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 19.96 on 98 degrees of freedom
Multiple R-squared:  0.04706,	Adjusted R-squared:  0.03734 
F-statistic:  4.84 on 1 and 98 DF,  p-value: 0.03016

> coef(summary(mod3))
             Estimate Std. Error   t value     Pr(>|t|)
(Intercept) 54.995466   2.822239 19.486464 1.755649e-35
treatment    8.780677   3.991249  2.199982 3.015810e-02
> alpha = 0.05
> bu1 <- 0
> for (i in 1:1000) {
+   outcome1 <- rnorm(50, mean=62.5, sd=20)
+ outcome2 <- rnorm(50, mean=57.5, sd=20)
+ outcome = c(outcome1, outcome2)
+ treatment <- c(rep.int( 1, 50), rep.int(0, 50))
+ 
+ y = data.frame(treatment, c(outcome1, outcome2))
+ mod3 = lm(outcome ~ treatment, dat=y)
+ summary(mod3)
+     i <- coef(summary(mod3))
+     p <- (pt(i[2,3], 99, lower.tail = FALSE, log.p = FALSE) *2)
+     if (p < alpha) 
+       bu1 = bu1 + 1
+   
+ }
> power=bu1/1000
> power
[1] 0.218

1. Find the power when the sample size is 1000 patients. (5 points)

```{r}
set.seed(63)
outcome1 <- rnorm(500, mean=62.5, sd=20)
outcome2 <- rnorm(500, mean=57.5, sd=20)
outcome = c(outcome1, outcome2)
treatment <- c(rep.int( 1, 500), rep.int(0, 500))

y = data.frame(treatment, c(outcome1, outcome2))
mod3 = lm(outcome ~ treatment, dat=y)
summary(mod3)
coef(summary(mod3))
alpha = 0.05
bu1 <- 0
for (i in 1:1000) {
  outcome1 <- rnorm(500, mean=62.5, sd=20)
outcome2 <- rnorm(500, mean=57.5, sd=20)
outcome = c(outcome1, outcome2)
treatment <- c(rep.int( 1, 500), rep.int(0, 500))

y = data.frame(treatment, c(outcome1, outcome2))
mod3 = lm(outcome ~ treatment, dat=y)
summary(mod3)
    i <- coef(summary(mod3))
    p <- (pt(i[2,3], 999, lower.tail = FALSE, log.p = FALSE) *2)
    if (p < alpha) 
      bu1 = bu1 + 1
  
}
power=bu1/1000
power

```
The test has a power of 97.7%

> set.seed(63)
> outcome1 <- rnorm(500, mean=62.5, sd=20)
> outcome2 <- rnorm(500, mean=57.5, sd=20)
> outcome = c(outcome1, outcome2)
> treatment <- c(rep.int( 1, 500), rep.int(0, 500))
> 
> y = data.frame(treatment, c(outcome1, outcome2))
> mod3 = lm(outcome ~ treatment, dat=y)
> summary(mod3)

Call:
lm(formula = outcome ~ treatment, data = y)

Residuals:
    Min      1Q  Median      3Q     Max 
-65.220 -13.672   0.617  14.399  61.529 

Coefficients:
            Estimate Std. Error t value Pr(>|t|)    
(Intercept)   59.223      0.915  64.721   <2e-16 ***
treatment      2.164      1.294   1.672   0.0948 .  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 20.46 on 998 degrees of freedom
Multiple R-squared:  0.002793,	Adjusted R-squared:  0.001794 
F-statistic: 2.796 on 1 and 998 DF,  p-value: 0.09483

> coef(summary(mod3))
             Estimate Std. Error   t value   Pr(>|t|)
(Intercept) 59.222552   0.915037 64.721484 0.00000000
treatment    2.163708   1.294058  1.672033 0.09483119
> alpha = 0.05
> bu1 <- 0
> for (i in 1:1000) {
+   outcome1 <- rnorm(500, mean=62.5, sd=20)
+ outcome2 <- rnorm(500, mean=57.5, sd=20)
+ outcome = c(outcome1, outcome2)
+ treatment <- c(rep.int( 1, 500), rep.int(0, 500))
+ 
+ y = data.frame(treatment, c(outcome1, outcome2))
+ mod3 = lm(outcome ~ treatment, dat=y)
+ summary(mod3)
+     i <- coef(summary(mod3))
+     p <- (pt(i[2,3], 999, lower.tail = FALSE, log.p = FALSE) *2)
+     if (p < alpha) 
+       bu1 = bu1 + 1
+   
+ }
> power=bu1/1000
> power
[1] 0.977

### Question 3 ###

**15 points**

Obtain a copy of the [football-values lecture](https://github.com/couthcommander/football-values).
Save the `2015/proj_rb15.csv` file in your working directory.  Read
in the data set and remove the first two columns.

1. Show the correlation matrix of this data set. (3 points)

```{r}
football = read.csv("proj_rb15.csv")
football$PlayerName <- NULL
football$Team <- NULL
c3 <- cor(football)

```
          rush_att  rush_yds  rush_tds   rec_att   rec_yds   rec_tds   fumbles
rush_att 1.0000000 0.9975511 0.9723599 0.7694384 0.7402687 0.5969159 0.8589364
rush_yds 0.9975511 1.0000000 0.9774974 0.7645768 0.7345496 0.6020994 0.8583243
rush_tds 0.9723599 0.9774974 1.0000000 0.7263519 0.6984860 0.5908348 0.8526904
rec_att  0.7694384 0.7645768 0.7263519 1.0000000 0.9944243 0.8384359 0.7459076
rec_yds  0.7402687 0.7345496 0.6984860 0.9944243 1.0000000 0.8518924 0.7224865
rec_tds  0.5969159 0.6020994 0.5908348 0.8384359 0.8518924 1.0000000 0.6055598
fumbles  0.8589364 0.8583243 0.8526904 0.7459076 0.7224865 0.6055598 1.0000000
fpts     0.9824135 0.9843044 0.9689472 0.8556928 0.8340195 0.7133908 0.8635550
              fpts
rush_att 0.9824135
rush_yds 0.9843044
rush_tds 0.9689472
rec_att  0.8556928
rec_yds  0.8340195
rec_tds  0.7133908
fumbles  0.8635550
fpts     1.0000000

1. Generate a data set with 30 rows that has a similar correlation
structure.  Repeat the procedure 10,000 times and return the mean
correlation matrix. (10 points)

```{r}
l<- f2 <- f3 <-f4 <- 0
for (l in 1:10) 
repeat 
  f2 <- replicate(8,sample(0:50,30,rep=TRUE))
f3=cor(f2)
f4=f3+f3
(l <- l +1)
if (l>9999) break


mean=f4/10000



```



1. Generate a data set with 30 rows that has the exact correlation
structure as the original data set. (2 points)

```{r}
repeat {
  c5 <- replicate(8, sample(0:1000, 30, rep=TRUE))
  c6 <- cor(c5)
  if (identical(c6,c3)) break
}
  


```


### Question 4 ###

**10 points**

Use \LaTeX to create the following expressions.

1. Hint: `\Rightarrow` (4 points)


\[
\P(B)= \sum_j P(B|A_j)P(A_j),

    \Rightarrow = P(A_i|B) =\frac {P (B|A_i)P(A_i)} {\sum_j (B\A_j)P(A_j)} 

\]





1. Hint: `\zeta` (3 points)

\[
\f\^(\zeta)=\int_{-infty}^\infty f(x) e^{-2\pi x\zeta} \, dx
\]   

1. Hint: `\partial` (3 points)

\[
\J= \frac {df} {dx} = \frac {\partial f} {\partial x_1} \ldots\frac {\partial f} {\partial x_1} = \begin{bmatrix}
\frac {\partial f_1} {\partial x_1} & \cdots & \frac {\partial f_1} {\partial x_n} \\ \vdots & \ddots & \vdots \\
\frac {\partial f_m} {\partial x_1}& \cdots & \frac {\partial f_m} {\partial x_n}


\end {bmatrix}
\]


### Notes/Grade

```{r}
# Question 3.2 (-8 pt)
library(MASS)
    corr.mat <- matrix(0, nrow=ncol(x), ncol=ncol(x))
    n <- 100
    for(i in seq(n)) {
      close <- mvrnorm(30, mu = colMeans(x), Sigma = var(x))
      corr.mat <- corr.mat + cor(close)/n
    }
corr.mat
# Question 3.3 (-2pt)

exact <- mvrnorm(30, mu = colMeans(x), Sigma = var(x), empirical = TRUE)
cor(exact)

```


40/50 points